DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2021-11-02 manuscript/manuscript.Rmd
abstract

abstract: |
  Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
  Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
  Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
  of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
  This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology. Despite these trends, accessibility for time data is not that good.
  R has become a widely used element of the biological data analyst toolkit. Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.

  Here we present `datelife`, a service implemented as an R package and a web site
  (www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.

  Main results:
    1. blah
    2. blah
    3. blah

  Results in a general context:
  All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis. Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
  How well this trees work for this purpose still needs to be tested.

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
wordcount         : "2650"

@@

documentclass     : "apa6"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    includes      :
      # in_header: preamble.tex
      after_body  : [figures.tex]

@@

The main goal of `datelife` is to generate a chronogram for any given combination of taxon names, based on expert scientific information.

>>>>

abstract: |
  Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
  Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
  Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
  of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
  This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology. Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.

  Here we present `datelife`, a service implemented as an R package and an Rshiny website application
  available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.

  The main workflow of `datelife` is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms. Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.

  Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology. How well this chronograms work for this purpose still needs to be tested.

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

  <!--  R has become mainstream in the biological-data analyst toolkit. Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow. -->

keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
wordcount         : "2650"

@@

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      after_body  : [figures.tex]

@@
